---
output: html_document
chapter: Inference
editor_options:
  chunk_output_type: console
---
# Inference
First, upload necessary package(s).
```{r}
library(tidyverse) #also uploads  dplyr
```
## 2.7 Exercises 

If you have not downloaded the data before,
```{r}
dir <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/"
filename <- "femaleControlsPopulation.csv"
url <- paste0(dir, filename)
x <- unlist(read.csv(url))
```

Or if you already have downloaded the data, then just upload it.

```{r}
x <- read.csv('femaleMiceWeights.csv')
x <- unlist(x)

# or 

x <- read.csv('femaleMiceWeights.csv') %>% unlist()

```

Check out what unlist does by typing ?unlist in the command. The second method is more concise because of the pipe %>%. 

### Question 1

<div class="question">
What is the average of these weights?
</div>

<div class="answer">

```{r}
mean(x)
```

### Question 2
<div class="question">
After setting the seed at 1, `set.seed(1)` take a random sample size 5. What is the absolute value (use `abs`) of the difference between the average of the sample and the average of all the values?
</div>

<div class="answer">

```{r}
set.seed(1)
avg_sample <- mean(sample(x,5)) # average of the sample of 5
avg_pop <- mean(x) # average of all values
abs(avg_sample - avg_pop) # absolute difference
```

### Question 3
<div class="question">
After setting the seed at 5, `set.seed(5)` take a random sample size 5. What is the absolute value (use `abs`) of the difference between the average of the sample and the average of all the values?
</div>

<div class="answer">

```{r}
set.seed(5)
avg_sample <- mean(sample(x,5)) # average of the sample of 5
avg_pop <- mean(x) # average of all values
abs(avg_sample - avg_pop) # absolute difference
```

### Question 4
<div class="question">
Why are the answers from 2 and 3 different?
</div>

<div class="answer">
```{r}
set.seed(2)
a <- sample(x,5)
a
```
```{r}
set.seed(3)
b <- sample(x,5)
b
```
```{r}
identical(a,b)
```
Notice that samples `a` and `b` differ. Since the seeds were different (`2` vs `3`), different random numbers were generated. Therefore, the answer is C: Because the average of the samples is a random variable.