---
output: html_document
chapter: Inference
editor_options:
  chunk_output_type: console
---
# Inference
First, upload necessary package(s).
```{r}
library(tidyverse) #also uploads  dplyr
```
## 2.7 Exercises 

If you have not downloaded the data before,
```{r}
dir <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/"
filename <- "femaleControlsPopulation.csv"
url <- paste0(dir, filename)
x <- unlist(read.csv(url))
```

Or if you already have downloaded the data, then just upload it.

```{r}
dat <- read.csv('femaleMiceWeights.csv')
bodyweight <- select(dat, Bodyweight)
x <- unlist(bodyweight)

# or use pipe %>%
x <- read.csv('femaleMiceWeights.csv') %>% select(Bodyweight) %>% unlist()

```

Check out what `unlist` does by typing `?unlist` in the command. The second method is more concise because of the pipe `%>%`, which allows multiple lines of commands to be in one continuous line. 

### Question 1

<div class="question">
What is the average of these weights?
</div>

<div class="answer">

```{r}
mean(x)
```

### Question 2
<div class="question">
After setting the seed at 1, `set.seed(1)` take a random sample size 5. What is the absolute value (use `abs`) of the difference between the average of the sample and the average of all the values?
</div>

<div class="answer">

```{r}
set.seed(1)
avg_sample <- mean(sample(x,5)) # average of the sample of 5
avg_pop <- mean(x) # average of all values
abs(avg_sample - avg_pop) # absolute difference
```

### Question 3
<div class="question">
After setting the seed at 5, `set.seed(5)` take a random sample size 5. What is the absolute value (use `abs`) of the difference between the average of the sample and the average of all the values?
</div>

<div class="answer">

```{r}
set.seed(5)
avg_sample <- mean(sample(x,5)) # average of the sample of 5
avg_pop <- mean(x) # average of all values
abs(avg_sample - avg_pop) # absolute difference
```

### Question 4
<div class="question">
Why are the answers from 2 and 3 different?
</div>

<div class="answer">
```{r}
set.seed(1) # question 2
a <- sample(x,5)
a
```
```{r}
set.seed(5) # question 3
b <- sample(x,5)
b
```
```{r}
identical(a,b) # these two samples are not identical
```
Notice that samples `a` and `b` differ. Since the seeds were different (`1` vs `5`), different random numbers were generated. Therefore, the answer is C: Because the average of the samples is a random variable.

### Question 5
<div class="question">
Set the seed at 1, then using a for-loop take a random sample of 5 mice in 1,000 times. Save these averages. What percent of these 1,000 averages are more than 1 gram away from the average of `x`?
</div>

<div class="answer">

```{r}
set.seed(1)
n <- 1000
res <- vector('double',n) 

for (i in seq(n)) {
  avg_sample <- mean(sample(x,5))
  res[[i]] <- avg_sample
}

mean(abs(res-mean(x)) > 1)
```
To make a for loop work in R, an empty vector needs to be created first. This can be achieved with the function `vector`. In this example, the empty vector is `res` (short for result). In the for loop, each average (`avg_sample`) from one repetition gets stored in `res`. 

### Question 6
<div class="question">
We are now going to increase the number of times we redo the sample from 1,000 to 10,000. Set the seed at 1, then using a for-loop take a random sample of 5 mice 10,000 times. Save these averages. What percent of these 10,000 averages are more than 1 gram away from the average of `x`?
</div>

<div class="answer">
```{r}
set.seed(1)
n <- 10000
res <- vector('double',n)
for (i in seq(n)) {
  avg_sample <- mean(sample(x,5))
  res[[i]] <- avg_sample
}
mean(abs(res-mean(x)) > 1)
```

### Question 7
<div class="question">
Note that the answers to 5 and 6 barely changed. This is expected. The way we think about the random value distributions is as the distribution of the list of values obtained if we repeated the experiment an infinite number of times. On a computer, we can't perform an infinite number of iterations so instead, for our examples, we consider 1,000 to be large enough, thus 10,000 is as well. Now if instead we change the sample size, then we change the random variable and thus its distribution.

Set the seed at 1, then using a for-loop take a random sample of 50 mice 1,000 times. Save these averages. What percent of these 1,000 averages are more than 1 gram away from the average of `x`?
</div>